A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-08-27 bioc-workflow.Rmd
Close #19: use markers rather than SVM.marker.set, and remove Golgi using fDataToUnknown

```{r env, echo=FALSE}
library("BiocStyle")
suppressPackageStartupMessages(library("knitr"))
suppressPackageStartupMessages(library("gridExtra"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRoloc"))
suppressPackageStartupMessages(library("pRolocdata"))
```

@@

@@

```{r plot2Drep, fig.width = 12}
par(mfrow = c(1, 2))
plot2D(hl[, hl$replicate == 1], fcol = "SVM.marker.set", main = "Replicate 1")
plot2D(hl[, hl$replicate == 2], fcol = "SVM.marker.set", main = "Replicate 2",
       mirrorX = TRUE)
```

@@

```{r plotPDres, fig.width=9}
## Re-order the colours for the phenoDisco output
cl <- getMarkerClasses(pdRes, "pd", verbose = FALSE)
cols <- getStockcol()[seq(cl)]
ind <- grep("Pheno", cl, invert = TRUE)
cols[ind] <- getStockcol()[seq(cl)][1:length(ind)]
cols[-ind] <- getStockcol()[seq(cl)][(length(ind) + 1):length(cl)]

@@

In the code chunk below we employ the use of a Support Vector Machine
(SVM) to learn a classifier on the labelled training data. The
training data is found in the `featureData` slot in the column called
`SVM.marker.set`. As previously mentioned one first needs to train the
classifiers parameters before an algorithm can be used to predict the
class labels of the proteins with unknown location. One of the most
common ways to optimise the parameters of a classifier is to partition
the labelled data in to training and testing subsets. In this
framework parameters are tested via a grid search using
cross-validation on the training partition. The best parameters chosen
from the cross-validation stage are then used to build a classifier to
predict the class labels of the protein profiles on the test
partition. Observed and expected classication results can be compared,
and then used to assess how well a given model works by getting an
estimate of the classiers ability to achieve a good generalisation
i.e. that is given an unknown example predict its class label with
high accuracy. In `pRoloc` algorithmic performance is estimated using
stratified 80/20 partitioning for the training/testing subsets
respectively, in conjuction with five-fold cross-validation in order
to optimise the free parameters via a grid search. This procedure is
usually repeated 100 times and then the best parameter(s) are selected
upon investigation of classifier accuracy, here we use the harmonic
mean of precision and recall; the macro F1 score. In the code chunk
below we demonstrate how to optimise the free parameters; `sigma` and
`cost`, of a classical SVM classifier with a Gaussian kernel using the
function `svmOptimisation`. As the number of labelled instances per
class varies from organelle to organelle, we can account for class
imbalance by setting specific class weights when generating the SVM
model. Below the weights, `w` are set to be inversely proportional to
the class frequencies.

```{r loadpdRes, include=FALSE}
pdRes <- hl 
load("Data/fusionparams.rda")
(w <- table(fData(pdRes)[, "SVM.marker.set"]))
w <- 1/w[names(w) != "unknown"]
```

```{r optimise, eval=FALSE}
(w <- table(fData(pdRes)[, "SVM.marker.set"]))
w <- 1/w[names(w) != "unknown"]

## 100 rounds of optimisation with five-fold cross-validation
params <- svmOptimisation(pdRes, fcol = "SVM.marker.set",
                          times = 100, xval = 5,
                          class.weights = w,
                          verbose = FALSE)
```

@@

We can use the function `svmClassification` to return a classification
result for all unlabelled instances in the dataset corresponding to
their most likely sub-cellular location. The algorithm parameters are
passed to the function, along with the class weights and the `fcol` to
tell the function where the labelled training data is located, in our
case, our marker proteins are located in the `featureData` with the
label `"SVM.marker.set"`.

```{r classify, eval=TRUE}
svmres <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "SVM.marker.set")
```

@@

```{r thresholds, fig.height=8, fig.width=8}
## First remove the markers
preds <- unknownMSnSet(svmres, fcol = "SVM.marker.set")

@@

```{r orgquants}
ts <- orgQuants(svmres, 
                fcol = "svm", 
                scol = "svm.scores",
                mcol = "SVM.marker.set", 
                t = .5)

preds <- getPredictions(svmres, 
                        fcol = "svm",
                        scol = "svm.scores", 
                        mcol = "SVM.marker.set",
                        t = ts)
```

@@

```{r classifyApp, eval=FALSE}
library("pRolocGUI")
pRolocVis(svmres, app = "classify", 
          fcol = "svm", 
	        scol = "svm.scores",
          mcol = "SVM.marker.set")
```

@@

The data is loaded and displayed on a PCA plot and a boxplot is used to
display the classifier scores by data class. On the left there is
a sidebar panel with sliders to control the thresholds upon which classifications 
are made. There are two types of cut-off that the user can choose from: (1)
"Quantile" and (2) "User-defined". By default, when the application is launched
quatile scoring is selected and set to 0.5, the median. The 
class-specific score thresholds that correspond to selecting the desired
quantile are shown on as red dots on the boxplot. The assignments on the PCA
plot are also updated according to the selected threshold. The quantile 
threshold can be set by moving the corresponding quantile slider. If one
wished to set their own cut-offs the "User-defined" radio button must be selected
and then the sliders for defining user-specified scores become active and the
scores and highlighted on the boxplot by blue dots. For more information we refer
users to the [`r Biocpkg("pRolocGUI")` tutorial vignette](http://bioconductor.org/packages/release/bioc/vignettes/pRolocGUI/inst/doc/pRolocGUI.html).

>>>>

```{r env, echo=FALSE, warning=FALSE}
library("BiocStyle")
suppressPackageStartupMessages(library("knitr"))
suppressPackageStartupMessages(library("gridExtra"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRoloc"))
suppressPackageStartupMessages(library("pRolocdata"))
```

@@

As demonstrated in [@hyper] and illustrated in the PCA plot above, the
Golgi apparatus proteins display a dynamic pattern, noting sets of
Golgi marker proteins that are distributed amongst other subcellular
structures, an observation supported by microscopy. As such, we are
going to reset the annotation of Golgi markers to unknown using the
`fDataTounknown` function. It is often used to replace empty strings
("") or missing values in the markers definition to a common
definition of *unknown localisation*.

```{r}
hl <- fDataToUnknown(hl, from = "Golgi", to = "unknown")
getMarkers(hl)
```

@@

```{r plot2Drep, fig.width = 12}
par(mfrow = c(1, 2))
plot2D(hl[, hl$replicate == 1], main = "Replicate 1")
plot2D(hl[, hl$replicate == 2], main = "Replicate 2",
       mirrorX = TRUE)
```

@@

```{r plotPDres, fig.width=9}
## Re-order the colours for the phenoDisco output
cl <- getMarkerClasses(pdRes, "pd")
cols <- getStockcol()[seq(cl)]
ind <- grep("Pheno", cl, invert = TRUE)
cols[ind] <- getStockcol()[seq(cl)][1:length(ind)]
cols[-ind] <- getStockcol()[seq(cl)][(length(ind) + 1):length(cl)]

@@

In the code chunk below we employ the use of a Support Vector Machine
(SVM) to learn a classifier on the labelled training data. As
previously mentioned one first needs to train the classifiers
parameters before an algorithm can be used to predict the class labels
of the proteins with unknown location. One of the most common ways to
optimise the parameters of a classifier is to partition the labelled
data in to training and testing subsets. In this framework parameters
are tested via a grid search using cross-validation on the training
partition. The best parameters chosen from the cross-validation stage
are then used to build a classifier to predict the class labels of the
protein profiles on the test partition. Observed and expected
classication results can be compared, and then used to assess how well
a given model works by getting an estimate of the classiers ability to
achieve a good generalisation i.e. that is given an unknown example
predict its class label with high accuracy. In `pRoloc` algorithmic
performance is estimated using stratified 80/20 partitioning for the
training/testing subsets respectively, in conjuction with five-fold
cross-validation in order to optimise the free parameters via a grid
search. This procedure is usually repeated 100 times and then the best
parameter(s) are selected upon investigation of classifier accuracy,
here we use the harmonic mean of precision and recall; the macro F1
score. In the code chunk below we demonstrate how to optimise the free
parameters; `sigma` and `cost`, of a classical SVM classifier with a
Gaussian kernel using the function `svmOptimisation`. As the number of
labelled instances per class varies from organelle to organelle, we
can account for class imbalance by setting specific class weights when
generating the SVM model. Below the weights, `w` are set to be
inversely proportional to the class frequencies.

```{r loadpdRes, include=FALSE}
pdRes <- hl 
load("Data/fusionparams.rda")
w <- table(getMarkers(pdRes, verbose = TRUE))
w <- 1/w[names(w) != "unknown"]
```

```{r optimise, eval=FALSE}
w <- table(getMarkers(pdRes, verbose = TRUE))
w <- 1/w[names(w) != "unknown"]

## 100 rounds of optimisation with five-fold cross-validation
params <- svmOptimisation(pdRes, fcol = "markers",
                          times = 100, xval = 5,
                          class.weights = w,
                          verbose = FALSE)
```

As mentioned previously, we reply on the default feature variable
`"markers"` to define the class labels and hence can ommit it. To use
another feature variables, one need to explicitly specify its name
using the `fcol` argument (for example `fcol = "markers2"`).

@@

We can use the function `svmClassification` to return a classification
result for all unlabelled instances in the dataset corresponding to
their most likely sub-cellular location. The algorithm parameters are
passed to the function, along with the class weights. As above, `fcol`
can be ignored as we use the labels defined in the default `"markers"`
feature variable.

```{r classify, eval=TRUE}
svmres <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "markers")
```

@@

```{r thresholds, fig.height=8, fig.width=8}
## First remove the markers
preds <- unknownMSnSet(svmres)

@@

```{r orgquants}
ts <- orgQuants(svmres, 
                fcol = "svm", 
                scol = "svm.scores",
                mcol = "markers", 
                t = .5)

preds <- getPredictions(svmres, 
                        fcol = "svm",
                        scol = "svm.scores", 
                        mcol = "markers",
                        t = ts)
```

@@

```{r classifyApp, eval=FALSE}
library("pRolocGUI")
pRolocVis(svmres, app = "classify",
          fcol = "svm",
	        scol = "svm.scores",
          mcol = "markers")
```

@@

The data is loaded and displayed on a PCA plot and a boxplot is used to
display the classifier scores by data class. On the left there is
a sidebar panel with sliders to control the thresholds upon which classifications 
are made. There are two types of cut-off that the user can choose from: (1)
"Quantile" and (2) "User-defined". By default, when the application is launched
quatile scoring is selected and set to 0.5, the median. The 
class-specific score thresholds that correspond to selecting the desired
quantile are shown on as red dots on the boxplot. The assignments on the PCA
plot are also updated according to the selected threshold. The quantile 
threshold can be set by moving the corresponding quantile slider. If one
wished to set their own cut-offs the "User-defined" radio button must be selected
and then the sliders for defining user-specified scores become active and the
scores and highlighted on the boxplot by blue dots. For more information we refer
users to the `r Biocpkg("pRolocGUI")` tutorial 
[vignette](http://bioconductor.org/packages/release/bioc/vignettes/pRolocGUI/inst/doc/pRolocGUI.html).
